---
title: "REVISION figures 2"
author: '`r Sys.info()["user"]`'
date: '`r format(Sys.time(), "%B %d, %Y %H:%M:%S %Z")`'
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
    highlight: pygments
    df_print: kable
params:
  sample_name: ~
  sample_path: ~
  out_rds_path: ~
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	dev = c("png", "jpeg", "pdf")
)
```




```{r lib}
library(hdf5r)
library(Seurat)
library(SoupX)
library(canceRbits)
library(dplyr)
library(patchwork)
library(DT)
library(Azimuth)
library(SCpubr)
library(tibble)
library(dittoSeq)
library(scRepertoire)
library(reshape2)
library(viridis)
library(grDevices)
library(ggpubr)
library(ggplot2)
library(data.table)
library(msigdbr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(tidyverse)
library(enrichplot)
library(RColorBrewer)
```



# Open corrected counts and annotated seurat object
Load the corrected (SoupX), normalized (SCTransformed) and annotated (Twice mapped - Tabula Sapiens Skin reference and PBMC azimuth reference) data.
```{r echo = TRUE, message=FALSE, warning=FALSE, out.width='100%'}
srat <- readRDS( "Ressler2024/metadata/100_SeurtObject.rds")

meta <- srat@meta.data
meta$WHO <- "SD"
meta$WHO[meta$patient %in% c("07", "08", "12", "17")] <- "CR"
meta$WHO[meta$patient %in% c("04", "06", "10", "11")] <- "PR"
srat <- AddMetaData(srat, meta$WHO, col.name = "WHO")
srat$WHO <- factor(srat$WHO, levels = c("CR", "PR", "SD"))
```

# Define levels and colors


```{r echo = TRUE, fig.height=25, fig.width=25, message=FALSE, warning=FALSE, out.width='100%'}

macro <- subset(srat, subset = anno_l1 == "Mono-Mac")

macro$seurat <- droplevels(macro$seurat_clusters)



colors_anno = c(
          "C1QC" = "#41ab5d",
          "SPP1" = "#e7298a" ,
           "FCN1" = "#fec44f",
         "Triple Neg." = "#41b6c4"
          
          
  
)


Idents(macro) <- "seurat_clusters"
cells <- WhichCells(macro, ident = c( "14") )
Idents(macro) <- "anno_l1"
macro <- SetIdent(macro, cells = cells, value = "SPP1")
macro$anno_l1 <- Idents(macro)

Idents(macro) <- "seurat_clusters"
cells <- WhichCells(macro, ident = c("8") )
Idents(macro) <- "anno_l1"
macro <- SetIdent(macro, cells = cells, value = "C1QC")
macro$anno_l1 <- Idents(macro)

Idents(macro) <- "seurat_clusters"
cells <- WhichCells(macro, ident = c("6") )
Idents(macro) <- "anno_l1"
macro <- SetIdent(macro, cells = cells, value = "FCN1")
macro$anno_l1 <- Idents(macro)

Idents(macro) <- "seurat_clusters"
cells <- WhichCells(macro, ident = c("20", "31") )
Idents(macro) <- "anno_l1"
macro <- SetIdent(macro, cells = cells, value = "Triple Neg.")
macro$anno_l1 <- Idents(macro)



macro@meta.data$anno_l1 <- factor(macro@meta.data$anno_l1, levels=c(
                                                                  "FCN1",
                                                                  "C1QC",
                                                                  "SPP1",
                                                                  "Triple Neg."
                                                                 ))

```


# Extended Data Figure 6A
```{r , eval=TRUE, fig.height=5, fig.width=5, message=FALSE, warning=FALSE,  out.width='100%'}

p <- SCpubr::do_DimPlot(sample = macro, 
                   group.by = "anno_l1", 
                   pt.size=0.5, 
                   colors.use = colors_anno,
                   label = TRUE, repel = TRUE, 
                   legend.position = "none",  
                   label.color = "black") + theme_minimal() + NoLegend() + theme(text = element_text(size=20))


p

DT::datatable(p$data, 
              caption = ("Extended Data Figure 6A"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))


```



# Extended Data Figure 6B - Buble plots markers subpopulations

```{r Fig2_bubblePlot_v1, eval=TRUE, fig.height=6, fig.width=10, message=FALSE, warning=FALSE,  out.width='100%'}


macro@meta.data$anno_l1 <- factor(macro@meta.data$anno_l1, levels=c(
                                                                  "Triple Neg.",
                                                                  "SPP1",
                                                                  "C1QC",
                                                                  "FCN1"
                                                                 ))


genes <- list(
               "FCN1" = c("FCN1"),
               "C1QC" = c("C1QC", "C1QB", "C1QA"),
               "SPP1" = c("SPP1"),
                "Triple Neg." = c("SPP1", "C1QC")
              )


p <- SCpubr::do_DotPlot(sample = macro,  
                        features = genes, group.by = "anno_l1",
                   font.size = 25, legend.length = 14,  legend.type = "colorbar", dot.scale = 8,  colors.use = c("#7fbc41","#b8e186", "#f7f7f7","#fde0ef", "#c51b7d"))

p

DT::datatable(p$data, 
              caption = ("Extended Data Figure 6B"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```

# Extended Data Figure 6C


```{r Fig3_BarPlot_Response, fig.height=6, fig.width=10, message=FALSE, warning=FALSE,  out.width='100%'}
macro$anno_l1 <- factor(macro$anno_l1, levels = c("Triple Neg.","SPP1",  "C1QC", "FCN1"))
q1 <- SCpubr::do_BarPlot( macro,
                         group.by = "anno_l1",
                         split.by = "WHO",
                        
                         position = "fill",
                   flip = TRUE,
                   order=FALSE,
                   legend.position = "bottom",
                   font.size = 30 ,  colors.use = colors_anno,
                   ylab = "Mono-Mac") + xlab("WHO \n response")




q1



DT::datatable(q1$data, 
              caption = ("Extended Data Figure 6C"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

```

# Extended Data Figure 6D

```{r fig.height=8, fig.width=15, message=FALSE, warning=FALSE, out.width='100%'}
macro$WHO <- factor(macro$WHO, levels = c("CR","PR","SD"))

df <- macro@meta.data %>%
    mutate_if(sapply(macro@meta.data, is.character), as.factor)  %>% 
    group_by( Pathological.Response,WHO, patient, anno_l1,    .drop = FALSE)%>% 
  summarise(Nb = n()) %>%
  mutate(C = sum(Nb)) %>%
  mutate(percent = Nb/C*100) %>%
  filter(percent != "NaN") %>%
  arrange(Pathological.Response, WHO,percent)




tmp <- df[df$anno_l1 == "FCN1", ]
p1 <- ggpaired(tmp, x = "WHO", y = "percent", id="patient", facet.by = "anno_l1", color = "black",fill = "WHO" , palette= c("white","yellow", "#fec44f"), group = "WHO",point.size = 2, xlab = "", ylab = "% among \n Mono-Mac", ncol = 13) + 
  stat_compare_means( method = "wilcox.test", paired = FALSE, label.y = c(75, 65, 65), hide.ns = FALSE, label = "p.signif",  bracket.size = 0.5, comparison = list(c("CR", "SD"), c("SD", "PR"), c("CR", "PR")), size = 8)  +
  theme(text = element_text(size = 24), axis.text.x = element_text(angle = 90)) + NoLegend()



tmp <- df[df$anno_l1 == "C1QC", ]
p2 <- ggpaired(tmp, x = "WHO", y = "percent", id="patient", facet.by = "anno_l1", color = "black",fill = "WHO" , palette= c("white","#b2df8a", "#33a02c"), group = "WHO",point.size = 2, xlab = "", ylab = "% among \n Mono-Mac", ncol = 13) + 
  stat_compare_means( method = "wilcox.test", paired = FALSE, label.y = c(75, 65, 65), hide.ns = FALSE, label = "p.signif",  bracket.size = 0.5, comparison = list(c("CR", "SD"), c("SD", "PR"), c("CR", "PR")), size = 8)  +
  theme(text = element_text(size = 24), axis.text.x = element_text(angle = 90)) + NoLegend()

 
tmp <- df[df$anno_l1 == "SPP1", ]
p3 <-  ggpaired(tmp, x = "WHO", y = "percent", id="patient", facet.by = "anno_l1", color = "black",fill = "WHO" , palette= c("white","#f1b6da", "#c51b7d"), group = "WHO",point.size = 2, xlab = "", ylab = "% among \n Mono-Mac", ncol = 13) + 
  stat_compare_means( method = "wilcox.test", paired = FALSE, label.y = c(75, 65, 65), hide.ns = FALSE, label = "p.signif",  bracket.size = 0.5, comparison = list(c("CR", "SD"), c("SD", "PR"), c("CR", "PR")), size = 8)  +
  theme(text = element_text(size = 24), axis.text.x = element_text(angle = 90)) + NoLegend()

 
tmp <- df[df$anno_l1 == "Triple Neg.", ]
p4 <- ggpaired(tmp, x = "WHO", y = "percent", id="patient", facet.by = "anno_l1", color = "black",fill = "WHO" , palette= c("white","#80cdc1", "#35978f"), group = "WHO",point.size = 2, xlab = "", ylab = "% among \n Mono-Mac", ncol = 13) + 
  stat_compare_means( method = "wilcox.test", paired = FALSE, label.y = c(75, 65, 65), hide.ns = FALSE, label = "p.signif",  bracket.size = 0.5, comparison = list(c("CR", "SD"), c("SD", "PR"), c("CR", "PR")), size = 8)  +
  theme(text = element_text(size = 24), axis.text.x = element_text(angle = 90)) + NoLegend()

 
p <- p1 + p2 + p3 + p4 + plot_layout(ncol = 4, guides = "collect") 

p



DT::datatable(rbind(p1$data, p2$data, p3$data, p4$data), 
              caption = ("Extended Data Figure 6D"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))


```


# Extended Data Figure 6E-H - Enrichment analysis



```{r  fig.height=12, fig.width=50, message=FALSE, warning=FALSE, include=TRUE, out.width='100%'}
# Find markers of main populations

gc()
options(future.globals.maxSize= 8*1024^3)
s.markers <- DElegate::FindAllMarkers2(macro, replicate_column = "patient", group_column = "anno_l1")

# universe
background <- rownames(macro)

ego_h <- list()
# pathways sets
GO_C5_gene_sets = msigdbr(species = "human", category = "C5", subcategory = "BP")
msigdbr_t2g_C5 = GO_C5_gene_sets %>% dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()

GO_H_gene_sets = msigdbr(species = "human", category = "H")
msigdbr_t2g_H = GO_H_gene_sets %>% dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()


CBioPlanet_gene_sets = read.csv2("CBioPlanet.csv", sep = ",")
CBioPlanet_gene_sets <- CBioPlanet_gene_sets[,c(2,4)]
colnames(CBioPlanet_gene_sets) <- c("gs_name", "gene_symbol")
CBioPlanet = CBioPlanet_gene_sets %>% dplyr::distinct(gs_name, gene_symbol) %>% as.data.frame()

# FNC1 
signature_h <- s.markers$feature[s.markers$group1 == "FCN1" & s.markers$padj < 0.05 & abs(s.markers$log_fc)>0.58]

##C5
ego_h[["FCN1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_C5)
b <- barplot(ego_h[["FCN1"]], title = "MSigDB C5 enrichments in FCN1", showCategory = 4 ) 
tmp1 <- b$data
tmp1$ID <- str_trunc(tmp1$ID, 150, "right")
tmp1$ID <- factor(tmp1$ID, levels = tmp1$ID[order(tmp1$Count)])
tmp1$type <- "FCN1"
##Hallmarks
ego_h[["FCN1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_H)
b <- barplot(ego_h[["FCN1"]], title = "MSigDB H enrichments in FCN1", showCategory = 3 ) 
tmp2 <- b$data
tmp2$ID <- str_trunc(tmp2$ID,150, "right")
tmp2$ID <- factor(tmp2$ID, levels = tmp2$ID[order(tmp2$Count)])
tmp2$type <- "FCN1"
##CBioPlanet
ego_h[["FCN1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = CBioPlanet)
b <- barplot(ego_h[["FCN1"]], title = "CBioPlanet enrichments in FCN1", showCategory = 4 ) 
tmp3 <- b$data
tmp3$ID <- str_trunc(tmp3$ID, 150, "right")
tmp3$ID <- factor(tmp3$ID, levels = tmp3$ID[order(tmp3$Count)])


p1 <- ggplot(tmp1, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["FCN1"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=20)) +
  ggtitle("FCN1") + 
  ylab("GO:BP")


p2 <- ggplot(tmp2, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["FCN1"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("") +
  ylab("Hallmarks")




# C1QC 
signature_h <- s.markers$feature[s.markers$group1 == "C1QC" & s.markers$padj < 0.05 & abs(s.markers$log_fc)>0.58]

##C5
ego_h[["C1QC"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_C5)
b <- barplot(ego_h[["C1QC"]], title = "MSigDB C5 enrichments in C1QC", showCategory = 4 ) 
tmp4 <- b$data
tmp4$ID <- str_trunc(tmp4$ID, 150, "right")
tmp4$ID <- factor(tmp4$ID, levels = tmp4$ID[order(tmp4$Count)])
tmp4$type <- "C1QC"
##Hallmarks
ego_h[["C1QC"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_H)
b <- barplot(ego_h[["C1QC"]], title = "MSigDB H enrichments in C1QC", showCategory = 3 ) 
tmp5 <- b$data
tmp5$ID <- str_trunc(tmp5$ID,150, "right")
tmp5$ID <- factor(tmp5$ID, levels = tmp5$ID[order(tmp5$Count)])
tmp5$type <- "C1QC"

##CBioPlanet
ego_h[["C1QC"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = CBioPlanet)
b <- barplot(ego_h[["C1QC"]], title = "CBioPlanet enrichments in C1QC", showCategory = 4 ) 
tmp6 <- b$data
tmp6$ID <- str_trunc(tmp6$ID, 150, "right")
tmp6$ID <- factor(tmp6$ID, levels = tmp6$ID[order(tmp6$Count)])
tmp6$type <- "C1QC"


p4 <- ggplot(tmp4, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["C1QC"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("C1QC") +
  ylab("")



p5 <- ggplot(tmp5, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["C1QC"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("") + 
  ylab("")



p6 <- ggplot(tmp6, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["C1QC"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("")  +
  ylab("")





# SPP1
signature_h <- s.markers$feature[s.markers$group1 == "SPP1" & s.markers$padj < 0.05 & abs(s.markers$log_fc)>0.58]

##C5
ego_h[["SPP1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_C5)
b <- barplot(ego_h[["SPP1"]], title = "MSigDB C5 enrichments in SPP1", showCategory = 4 ) 
tmp7 <- b$data
tmp7$ID <- str_trunc(tmp7$ID, 150, "right")
tmp7$ID <- factor(tmp7$ID, levels = tmp7$ID[order(tmp7$Count)])
tmp7$type <- "SPP1"

##Hallmarks
ego_h[["SPP1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = msigdbr_t2g_H)
b <- barplot(ego_h[["SPP1"]], title = "MSigDB H enrichments in SPP1", showCategory = 3 ) 
tmp8 <- b$data
tmp8$ID <- str_trunc(tmp8$ID,150, "right")
tmp8$ID <- factor(tmp8$ID, levels = tmp8$ID[order(tmp8$Count)])
tmp8$type <- "SPP1"

##CBioPlanet
ego_h[["SPP1"]] <- enricher(gene = signature_h, universe = background, TERM2GENE = CBioPlanet)
b <- barplot(ego_h[["SPP1"]], title = "CBioPlanet enrichments in SPP1", showCategory = 4 ) 
tmp9 <- b$data
tmp9$ID <- str_trunc(tmp9$ID, 150, "right")
tmp9$ID <- factor(tmp9$ID, levels = tmp9$ID[order(tmp9$Count)])
tmp9$type <- "SPP1"


p7 <- ggplot(tmp7, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["SPP1"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("SPP1") +
  ylab("")



p8 <- ggplot(tmp8, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["SPP1"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("") +
  ylab("")



p9 <- ggplot(tmp9, aes(Count, ID)) +
    geom_bar(stat = "identity", color="black", fill =  colors_anno["SPP1"]) +
   theme_classic()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "black",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18)) +
  ggtitle("") +
  ylab("")

p3 <- ggplot(tmp9, aes(Count, ID)) +
    geom_bar(stat = "identity", color="white", fill =  "white") +
   theme_minimal()+ 
    geom_text(
        aes(label = (paste( "padj=",  format(p.adjust,scientific = TRUE, digits = 2)))),
        color = "white",
        size = 6,
        hjust=1,
        position = position_dodge(0.5)  ) +
     theme(text = element_text(size=18), axis.text = element_text(color = "white")) +
  ggtitle("") +  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
    ylab("BioPlanet") + xlab ("")




(p1 + p4 + p7) + plot_layout(ncol=3)
DT::datatable(rbind(p1$data, p4$data, p7$data), 
              caption = ("Extended Data Figure 6E"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))


(p2 + p5 + p8) + plot_layout(ncol=3)
DT::datatable(rbind(p2$data, p5$data, p8$data), 
              caption = ("Extended Data Figure 6F"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))

(p3 + p6 + p9) + plot_layout(ncol=3)
DT::datatable(rbind(p3$data, p6$data, p9$data), 
              caption = ("Extended Data Figure 6G"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```



Answer to reviewer 3

```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

IFN_TAM <- list(c("CASP1","CASP4", "CCL2","CCL3","CCL4","CCL7","CCL8", "CD274","CD40","CXCL2","CXCL3","CXCL9","CXCL10","CXCL11", "IDO1","IFI6","IFIT1","IFIT2","IFIT3","IFITM1","IFITM3","IRF1","IRF7","ISG15","LAMP3","PDCD1LG2","TNFSF10","C1QA" ,"C1QC","CD38","IL4I1","ISG15","TNFSF10","IFI44L"))

macro <- AddModuleScore(macro,features = IFN_TAM, name = "IFN_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "IFN_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Inflam_TAM <- list(c("CCL2","CCL3","CCL4","CCL5","CCL20","CCL3L1","CCL3L3","CCL4L2","CCL4L4","CXCL1","CXCL2","CXCL3","CXCL5","CXCL8","G0S2","IL1B","IL1RN","IL6","INHBA","KLF2", "KLF6","NEDD9","PMAIP1","S100A8","S100A9","SPP1"))

macro <- AddModuleScore(macro,features = Inflam_TAM, name = "Inflam_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Inflam_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

LA_TAM <- list(c("ACP5","AOPE","APOC1","ATF1","C1QA", "C1QB", "C1QC","CCL18","CD163","CD36","CD63","CHI3L1","CTSB","CTSD", "CTSL","F13A1","FABP5","FOLR2","GPNMB","IRF3","LGALS3","LIPA","LPL","MACRO","MerTK","MMP7","MMP9", "MMP12","MRC1","NR1H3","NRF1","NUPR1","PLA2G7","RNASE1","SPARC","SPP1","TFDP2","TREM2","ZEB1"))

macro <- AddModuleScore(macro,features = LA_TAM, name = "LA_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "LA_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Angio_TAM <- list(c("ADAM8","AREG","BNIP3","CCL2","CCL4","CCL20","CD163","CD300E","CD44","CD55","CEBPB","CLEC5A","CTSB","EREG","FCN1","FLT1","FN1","HES1","IL1B","IL1RN","CXCL8","MAF","MIF","NR1H3","OLR1","PPARG","S100A8","S100A9","S100A12","SERPINB2","SLC2A1","SPIC","SPP1","THBS1","TIMP1","VCAN","VEGFA"))

macro <- AddModuleScore(macro,features = Angio_TAM, name = "Angio_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Angio_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```

```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Reg_TAM <- list(c("CCL2","CD274","CD40","CD80","CD86","CHIT1","CX3CR1","HLA-A","HLA-C","HLA-DQA1","HLA-DQB1","HLA-DRA","HLA-DB1", "HLA-DB5","ICOSLG","IL10","ITGA4","LGALS9","MACRO","MRC1","TGFB2"))

macro <- AddModuleScore(macro,features = Reg_TAM, name = "Reg_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Reg_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Prolif_TAM <- list(c("CCNA2","CDC45","CDK1","H2AFC","HIST1H4C","HMGB1","HMGN2","MKI67","RRM2","STMN1","TOP2A","TUBA1B","TUBB","TYMS"))

macro <- AddModuleScore(macro,features = Prolif_TAM, name = "Prolif_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Prolif_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```

```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

RTM_TAM <- list(c("C1QA","C1QB","C1QC","CCL7","CD163","CD5L","CD74","CETP","FOLR2","HLA-DPA1", "HLA-DPB1","HLA-DRB1","MACRO","MAF","MS4A7","SLC40A1","VCAM1","VSIG4"))

macro <- AddModuleScore(macro,features = RTM_TAM, name = "RTM_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "RTM_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

MT_RTM_TAM <- list(c("MT1E","MT1F","MT1G","MT1H","MT1M","MT1X", "MT2A","MTG1E","MTG1M","CXCL9", "CXCL10","GBP1","IFITM3","LGALS2","PARP14","SLC30A1","SLC39A8","STAT1","TNFSF13B"))

macro <- AddModuleScore(macro,features = MT_RTM_TAM, name = "MT_RTM_TAM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "MT_RTM_TAM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Classical_TIM <- list(c("S100A4","S100A6","S100A8","S100A9","S100A12","S100A13","AREG","CCL4","CCR2","CD14","CD300E","CD36","CEBPD","CLEC11A","CLEC12A","CSF1R","CTSS","CX3CR1","CXCL2","CXCR4","EGR1","FCN1","FOS","FTH1","IL1B","ITGAM","JUNB","LGALS2","LY6C","LYZ","MAF","MAFB","MPEG1","NFKB1","NFKBIA","NLRP3","NR4A1","NR4A2","OSM","PTGS2","RGS2","SELL","THBS1","VCAN"))

macro <- AddModuleScore(macro,features = Classical_TIM, name = "Classical_TIM", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Classical_TIM1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

NonClassical_Monocytes <- list(c("CX3CR1","CCL5","CD52","CDH23","CDKN1C","FCGR3A", "FCGR3B","FCGR4","GNLY","GZMB","ICAM2","IFITM1","IL32","KLRC4","KLRK1","LILRA", "LILRB1","MS4A7","MTSS1","PRF1","SLC44A2"))

macro <- AddModuleScore(macro,features = NonClassical_Monocytes, name = "NonClassical_Monocytes", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "NonClassical_Monocytes1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```


```{r eval=TRUE, fig.height=8, fig.width=20, message=FALSE, warning=FALSE,  out.width='100%'}

Intermediate_Monocytes <- list(c("CCL2","CCL8","CX3CR1","CXCL10","FCGR3A","IFITM2","IL1RN","IRF7","ISG15","ISG20","LILRB1","LILRA1","MS4A7","MTSS1","RHOC","SERPINA1","SIGLEC10","STAT1","TCF7L2","TNFSF10","TNFSF13B"))

macro <- AddModuleScore(macro,features = Intermediate_Monocytes, name = "Intermediate_Monocytes", nbin = 5 )

  p <- SCpubr::do_FeaturePlot(macro, features = "Intermediate_Monocytes1", pt.size = 0.2, legend.length =  8)
  DT::datatable(p$data, 
              caption = ("Extended Data Figure 6J"),
              extensions = 'Buttons', 
              options = list( dom = 'Bfrtip',
              buttons = c( 'csv', 'excel')))
```



```{r eval=TRUE, fig.height=30, fig.width=10, message=FALSE, warning=FALSE,  out.width='100%'}

v1 <- Seurat::VlnPlot(macro, 
                     features = "IFN_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend() 

v2 <- Seurat::VlnPlot(macro, 
                     features = "Inflam_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v3 <- Seurat::VlnPlot(macro, 
                     features = "LA_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v4 <- Seurat::VlnPlot(macro, 
                     features = "Angio_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v5 <- Seurat::VlnPlot(macro, 
                     features = "Reg_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v6 <- Seurat::VlnPlot(macro, 
                     features = "Prolif_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v7 <- Seurat::VlnPlot(macro, 
                     features = "RTM_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v8 <- Seurat::VlnPlot(macro, 
                     features = "MT_RTM_TAM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v9 <- Seurat::VlnPlot(macro, 
                     features = "Classical_TIM1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v10 <- Seurat::VlnPlot(macro, 
                     features = "NonClassical_Monocytes1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()
v11 <- Seurat::VlnPlot(macro, 
                     features = "Intermediate_Monocytes1", 
                     group.by = "seurat_clusters", 
                     pt.size = 0  
                     ) & NoLegend()

v1 + v2 +v3+v4+v5+v6+v7+v8+v9+v10+v11
```



```{r eval=TRUE, fig.height=25, fig.width=5, message=FALSE, warning=FALSE,  out.width='100%'}

v1 <- Seurat::VlnPlot(macro, 
                     features = "IFN_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0  ,
                    cols = colors_anno,
                     ) & NoLegend() 

v2 <- Seurat::VlnPlot(macro, 
                     features = "Inflam_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v3 <- Seurat::VlnPlot(macro, 
                     features = "LA_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v4 <- Seurat::VlnPlot(macro, 
                     features = "Angio_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v5 <- Seurat::VlnPlot(macro, 
                     features = "Reg_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v6 <- Seurat::VlnPlot(macro, 
                     features = "Prolif_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v7 <- Seurat::VlnPlot(macro, 
                     features = "RTM_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v8 <- Seurat::VlnPlot(macro, 
                     features = "MT_RTM_TAM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v9 <- Seurat::VlnPlot(macro, 
                     features = "Classical_TIM1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v10 <- Seurat::VlnPlot(macro, 
                     features = "NonClassical_Monocytes1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()
v11 <- Seurat::VlnPlot(macro, 
                     features = "Intermediate_Monocytes1", 
                     group.by = "anno_l1", 
                     pt.size = 0    ,
                    cols = colors_anno,
                     ) & NoLegend()

v1 + v2 +v3+v4+v5+v6+v7+v8+v9+v10+v11
```
